---
title: "Genomic selection with breedR: ssGBLUP, GWAS and genomic prediction"
author: "breedR"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Genomic selection with breedR: ssGBLUP, GWAS and genomic prediction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# The code in this vignette drives the BLUPF90 backend, which is downloaded at
# install time and is not available when the vignette is built (e.g. on CRAN).
# Chunks are therefore shown but not evaluated; illustrative output is described
# in the text. Run the code in an R session with the binaries installed.
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

## Overview

This vignette walks through the single-step genomic pipeline in breedR, end to
end:

1. prepare genotypes,
2. fit a single-step GBLUP (ssGBLUP) model,
3. run a genome-wide association study (ssGWAS),
4. predict genomic values for new animals,

and points to the Bayesian (Gibbs) and RENUMF90 workflows. Every step wraps a
program from the [BLUPF90 family](https://nce.ads.uga.edu/wiki/), so the backend
must be installed first.

## Installing the backend

The core `blupf90+` binary is downloaded when the package is installed. The
genomic programs and `renumf90` are installed separately:

```{r install-binaries}
library(breedR)

install_genomic_programs()   # preGSf90, postGSf90, predf90, gibbsf90+, ...
install_renumf90()           # renumf90
```

Throughout, we use the built-in `globulus` dataset (a forest genetic trial) and
a small synthetic set of markers so the workflow is self-contained.

```{r data}
data(globulus)
str(globulus[, 1:7])
```

## 1. Preparing genotypes

Genotypes are written in the BLUPF90 format with `write_snp_file()`: one row per
animal, the animal id followed by a fixed-width string of `0/1/2` codes
(`5` = missing). The animal ids **must** match the ids used in the pedigree /
data (breedR reconciles them internally through a cross-reference file).

```{r genotypes}
# A synthetic 200-marker panel for the first 60 animals in the trial.
set.seed(1)
gen_ids <- globulus$self[1:60]
G <- matrix(sample(0:2, length(gen_ids) * 200, replace = TRUE,
                   prob = c(0.25, 0.5, 0.25)),
            nrow = length(gen_ids))

write_snp_file(G, ids = gen_ids, file = "genotypes.txt")

# Fractional (imputed) dosages are also supported:
# write_snp_file(dosages, ids = gen_ids, file = "genotypes.txt", fractional = TRUE)
```

A map file (`SNP_ID CHR POS`) is optional; supply it to get chromosome and
position columns in the GWAS output.

## 2. Fitting a single-step GBLUP model

Pass a `genomic` list to `remlf90()`. breedR runs PREGSF90 first (genotype QC and
G-matrix construction), builds the required cross-reference file from its own
pedigree renumbering, and then fits the model with `blupf90+` using the combined
pedigree-genomic relationship matrix.

Set `save_ginverse = TRUE` if you intend to run a GWAS afterwards (see step 3);
it costs some extra I/O, so leave it off for a plain evaluation.

```{r ssgblup}
res <- remlf90(
  fixed   = phe_X ~ gg,
  genetic = list(model    = 'add_animal',
                 pedigree = globulus[, 1:3],
                 id       = 'self'),
  data    = globulus,
  genomic = list(snp_file      = "genotypes.txt",
                 # map_file    = "map.txt",   # optional, for GWAS positions
                 minfreq       = 0.05,        # MAF filter
                 save_ginverse = TRUE)        # needed for the GWAS below
)

summary(res)
```

The genotype QC report is attached to the fitted object:

```{r ssgblup-qc}
res$genomic$n_snp_total    # markers read
res$genomic$n_snp_passed   # markers passing QC
res$genomic$freq           # allele frequencies after cleaning
```

For a clean synthetic panel like the one above you should see all 200 markers
passing QC, and `summary(res)` reporting the variance components and estimated
heritability just as for a pedigree-only animal model.

## 3. Genome-wide association (ssGWAS)

`postgsf90()` back-solves the genomic breeding values into SNP effects. It reads
the genomic G-inverse saved by the fit, so the model **must** have been fitted
with `save_ginverse = TRUE` (otherwise `postgsf90()` stops and tells you to
refit). Run it in the same R session as the fit: it reads that fit's working
directory, `res$reml$dir`, which lives under `tempdir()` and goes away with the
session. Each fit has its own, so fitting another model in between is harmless.

```{r gwas}
gwas <- postgsf90(res,
                  windows_variance = 20,     # variance explained per 20-SNP window
                  manhattan_plot   = TRUE)

head(gwas$snp_sol)    # per-SNP solutions, weights and variances
head(gwas$manhattan)  # one row per marker: effect, position (if a map was given)
head(gwas$windows)    # variance explained by genomic windows
```

`gwas$snp_sol` has one row per marker; with a `map_file` supplied at fit time,
`gwas$manhattan` carries the chromosome and position needed to draw a Manhattan
plot.

## 4. Predicting genomic values for new animals

`predf90()` predicts direct genomic values (DGV) for genotyped animals that were
not in the evaluation, using the SNP effects `postgsf90()` just wrote. Run it in
the same session, and pass it the directory `postgsf90()` returns — that is
where the SNP effects are.

```{r prediction}
# genotypes for the animals to predict, in the same BLUPF90 format
predictions <- predf90(snp_file   = "new_genotypes.txt",
                       dir        = gwas$dir, # where postgsf90() wrote snp_pred
                       use_mu_hat = TRUE)   # add the base so DGV are on the GEBV scale

head(predictions)   # data.frame(id, call_rate, dgv)
```

Reliabilities (`acc = TRUE`) additionally require an `OPTION snp_var` file from
the GWAS run; see the package README "Known limitations".

## Bayesian inference and other workflows

**Gibbs sampling.** For Bayesian variance-component estimation (and
threshold/categorical traits) use `gibbsf90()` in place of `remlf90()`:

```{r gibbs}
res_b <- gibbsf90(phe_X ~ gg,
                  genetic   = list(model = 'add_animal',
                                   pedigree = globulus[, 1:3], id = 'self'),
                  data      = globulus,
                  n_samples = 50000, burnin = 5000, thin = 10)

colMeans(res_b$samples)              # posterior means of the variance components
diag <- postgibbsf90(res_b)          # convergence diagnostics (Geweke, ESS, HPD)
```

**RENUMF90 data preparation.** For alphanumeric ids, unknown-parent groups or
multi-trait models with per-trait effects, prepare the data with
`renumf90_from_data()` (which takes R data.frames directly) and fit with
`remlf90_from_renum()`.

**Prediction validation.** `validate_prediction()` implements the LR validation
method. Note the current backend limitation described in the package README
("Known limitations").

## See also

- `?remlf90` — the main fitting interface and the full `genomic` list.
- `?postgsf90`, `?predf90`, `?gibbsf90` — the genomic wrappers.
- `inst/doc/pregsf90_postgsf90_options.md` — the full PREGSF90/PostGSF90 option
  reference and how the `genomic` list maps to it.
